Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2023-06-09 noise_covid_v2.Rmd
update figures

Core challenges in epidemiology include estimating the efficacy of control measures, and understanding the impact of events of interest on incidence case counts. Allocating healthcare resources toward individuals with higher-than-average transmission risk can disproportionately reduce the population-level effective reproduction number (i.e., the average number of secondary infections caused by a randomly selected individual; Althouse et al. 2020). In other words, leveraging transmission variability among individuals to target interventions can be a highly effective strategy for reducing population spread. However, implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable. Conversely, evaluating control strategies using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control. The impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy. Similarly, the impacts of events involving congregation (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes or even mean changes per county.

@@

Since evaluating control strategies using only course-scale changes in incidence over time may fail to account for parts of the context of transmission and control, we should aim to learn about individual-level heterogeneity as a means to understand control strategies more comprehensively. Similarly, we should aim to learn about individual-level heterogenity following congregation events, as estimates of individual-level variability are used to identify so-called "superspreading" dynamics: high levels of individual-level variability indicate that some individuals cause more onward transmission (via secondary infections) than others, and this in turn scales up to influence resulting incidence time series following these events.

>>>>

Evaluating infectious disease dynamics using only course-scale changes in incidence over time may fail to account for important parts of the social-biological context of transmission and control. For instance, the impact of non-pharmateutical interventions (NPIs; e.g., mask mandates) is often assessed in terms of changes in *mean* incidence, yet mask mandates are implemented and adhered to heterogeneously, so their effects may be context-dependent, leading to geographic and temporal variation in impact for the same policy. Similarly, the impacts of events involving congregation (such as national holidays) may result in changes in epidemic trajectories that differ based on context, and these changes may not be elucidated by simply examining aggregate mean changes or even mean changes per county.

Conversely, implementing and evaluating targeted control strategies has typically required detailed data on individual-level variation in transmission rates, which are often unavailable. We present a method to identify shifts in population-level dispersion in incidence, so the impact of events on superspreading dynamics can be assessed using freely available population-level data.

@@
